Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00

### 2019-01-28 manuscript.Rmd
text updates, move sup figs around

## RNA-seq is the predominant technology for deciphering transcriptomes
From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid. The differentiation, development, and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, also known as the transcriptome. Today, RNA-sequencing (RNA-seq) is the predominant technology for quantifying the transcriptome. Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function. To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids. These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].

@@

# Methods
## Identifying potential eye samples with exhaustive search of SRA
We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a]. We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, iPSC derived tissue, organoids, and immortalized cell lines. In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project. Using SRA metadata from each study we extracted sample and run accessions, library type, tissue of origin, and sub-tissue of origin. Any of the preceding information missing from the SRA metadata was added by hand, when available. Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.

@@

## EiaD 2019 demonstrates that organoids, iPSC-derived, and fetal tissue retina have closely related transcriptomes
There are, currently, three approaches to studying developing human retina: 1. post-mortem fteal tissue, 2. organoids, and 3. iPSC-derived cells. As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level. Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.

To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`). Here we see three distinct groupings: adult retina (2), developing retina (3), and iPSC/undifferentiated (1). We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X'). All of the iPSC-based retina samples are older than 50 days.

```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F, warning = F}
tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, iPSC retina, and fetal retina tissue have highly similar transcriptomes. The zoom inset shows the retina samples. The "Sub-Tissue Cluster" shading shows the cluster membership of the three major groups. The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.')

@@

EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues. Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression. Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC and organoid based developmental methods to be developed. Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina developmental .

>>>>

## RNA-seq is the predominant technology for deciphering transcriptomes
From anterior to posterior, the human eye is composed of the cornea, lens, retina, retinal pigment epithelium (RPE), and choroid. The differentiation, development, and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, also known as the transcriptome. Today, RNA-sequencing (RNA-seq) is the predominant technology for quantifying the transcriptome. Analysis of the transcriptome across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function. To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), differentiated stem cells, immortalized cell lines, and most recently, organoids. These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014b; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014b; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015a; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].

@@

# Methods
## Identifying potential eye samples with exhaustive search of SRA
We searched the SRA with the SRAdb R package for eye related tissues using the query ‘cornea|retina|RPE|macula|fovea|choroid|sclera|iris|lens|eye’ across all columns and rows in the 'sra' table [@leinonenSequenceReadArchive2011; @zhuSRAdbQueryUse2013a]. We hand selected relevant studies and selected healthy, control or unmodified samples spanning primary adult tissue, primary fetal tissue, stem cell derived tissue, organoids, and immortalized cell lines. In order to compare gene expression in the eye against expression in other body tissues, we obtained samples from `r core_tight_2019 %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% unique() %>% length()` different body tissues that contained at least 10 male and 10 female samples from the GTEx project. Using SRA metadata from each study we extracted sample and run accessions, library type, tissue of origin, and sub-tissue of origin. Any of the preceding information missing from the SRA metadata was added by hand, when available. Stem cell derived tissues and cell lines are marked as sub-tissues of the tissue they model.

@@

## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
There are, currently, three approaches to studying developing human retina: 1. post-mortem fteal tissue, 2. organoids, and 3. stem cell-derived cells. As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level. Should organoid and stem cell-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.

To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`). Here we see three distinct groupings: adult retina (2), developing retina (3), and stem cell/undifferentiated (1). We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X'). All of the stem cell-based retina samples are older than 50 days.

```{r tsneFetalRetina, fig.height = 4, fig.cap=tsneFetalRetina_cap, echo=F, warning = F}
tsneFetalRetina_cap <- fig_cap(name='tsneFetalRetina', caption = 'Organoid retina, stem cell retina, and fetal retina tissue have highly similar transcriptomes. The zoom inset shows the retina samples. The "Sub-Tissue Cluster" shading shows the cluster membership of the three major groups. The shapes of the points show the different origin types - notable types include the square for adult, the \'X\' for organoid under 30 days of differentiation, and the diamond for organoid over 30 days of differentiation.')

@@

EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues. Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression. Several human and non-human primate studies have been posted in the past year on the bioRxiv pre-print server. As the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for stem cell and organoid based developmental methods to be developed. Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina developmental .
